Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-04-17 paperDistance/paper.tex
paper stuff

We would like to infer the true parallax $\varpi_{\true}$ for
each star given its parallax measurement $\varpi$ and associated
uncertainty $\sigma_{\varpi}$ from \tgas.
We assume (as \gaia\ suggests) that the likelihood is Gaussian
\begin{equation}
p(\varpi \given \varpi_{\true}, I) = \mathcal{N}(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \quad ,
\label{eq:gaialike}
\end{equation}
where $I$ is a placeholder containing our assumptions, including all
uncertainty estimates, and $\mathcal{N}(x \given \mu, \sigma^2)$ is a
Gaussian function for $x$ with mean $\mu$ and variance $\sigma^2$.
This is the probability of the data given the true parallax, but we
want to flip it around and infer the true parallax, or distance, to
the star given the measured parallax and it's uncertainty. We do this
using probability theory
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.
The strength of our inference method here lies in how we generate our prior
pdf for the parallax $p(\varpi_{\true})$.
Instead of assuming a functional form for the spatial distribution of
stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar
models (\citealt{gaia16}), we build a noise-deconvolved estimate of the color--magnitude
diagram (\cmd) from the observed colors and absolute magnitudes of our
full set of stars.
This \cmd\ model becomes, for each star---given its observed color and
magnitude (and uncertainties on those)---a prior pdf for the distance
to that star.

That said, this is \emph{not} exactly how we do the inference here.
Here we perform the inference in a two-dimensional space that is a
transform of the \cmd....[HOGG: Syntactical marker]

To generate our \cmd\ prior we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. The data are a transformation of the dust corrected absolute magnitude, and the dust corrected colors.

@@

The transformation of the dust corrected absolute magnitude, Equation \ref{eq:transform}, is required to to keep the uncertainties gaussian, which, under some assumptions along with this transformation, we maintain. Gaussian uncertainties are required to use \xd\, more about this and our dust corrections below.

Our true, latent values are the true $J-K$ color $C_n$ and the true $J$ band absolute magnitude $M_{J,n}$ for each star, which we encapsulate in the true vector $\vec{Y}$. The observed data is the dust corrected $J-K$ color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\vec{y}$. Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the Greene dust map

\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\vec{Y} = [10^{0.2\,M_{J,n} + 2}, C_n] \\
\mathrm{data \, vector} \quad &\vec{y} = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{assumptions} \quad &I = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}

@@

Working with these variables, we then have a likelihood for the data, encapsulating both the \gaia\ parallaxes and \tmass\ photometry of

\begin{equation}
p(\vec{y} \given \vec{Y}) = \mathcal{N}(\vec{y} \given \vec{Y}, \sigma_{\vec{y}})
\label{eq:fulllike}
\end{equation}

To generate our prior, we fit the data in the slightly transformed \cmd\ space, Equation \ref{eq:transform}, using \xd\ which deconvolves the data with its uncertainties, which are heterogeneous measurement uncertainties. It generates the denoised, underlying distribution from which the uncertain data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the marginalized likelihood of the data, marginalized over the true data.

@@

\begin{eqnarray}
\alpha &=& \left\{\mathrm{amp}_k, \mu_k, V_k\right\}_{k=1}^K \; k \in \mathrm{Gaussians} \nonumber\\
       &=& \argmax_{\alpha} \, \Pi_n \, p(\vec{y}_n \given \alpha) \nonumber\\
       &=& \argmax_{\alpha} \, \Pi_n \, \int p(\vec{y}_n \given \vec{Y}_n) \, p(\vec{Y} \given \alpha)\mathrm{d\vec{Y}}
\label{eq:xdmml}
\end{eqnarray}

To infer $\varpi_{\true}$ from the likelihood of the \gaia\ data, Equation \ref{eq:gaialike}, we first do the inference in the 2D space of the \cmd

\begin{equation}
p(\vec{Y} \given \vec{y}, \alpha) = \frac{1}{Z} \, p(\vec{y} \given \vec{Y}) \, p(\vec{Y} \given \alpha)
\label{eq:posterior}
\end{equation}

We then project the 2D posterior into the 1D absolute magnitude/parallax space

\begin{eqnarray}
p(\vec{Y}_0 \given \vec{y}, \alpha) &\rightarrow& p(\varpi_{\true} \given y, \alpha) \, f(\varpi_{\true}) \\
f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}

And with a jacobian this becomes a posterior on distance to the star if you wish.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{priorNgaussComparison.png}
\caption{Prior: the XD generated mixture of gaussians we use as our prior. To determine the number of gaussians we used cross validation, using 70\% of our sample as a training set and 30\% as the test set. Increasing the number of gaussians we used continued to increase the likelihood of the test set of data, showing that we are no over fitting. Using more gaussians as our prior did not substantially change our posterior results, but it did start to highlight the features of the CMD.}
\label{fig:prior}
\end{figure}

@@

Long discussion of dust.

>>>>

\subsection{Setting up the Inference}
Think about presentation, prioritizing/highlighting prior but logically building it from an full inference perspective and needing a prior, then dive into prior stuff. It's confusing right now jumping between inference and prior building !! 
\subsection{Generating the Prior}

Under these assumptions, we would like to infer the true parallax $\varpi_{\true}$ for each star given its measured parallax $\varpi$ and associated uncertainty $\sigma_{\varpi}$ from \tgas.
The strength of our inference method here lies in how we generate our prior information on the true parallax.
Supplementary to this trigonometric measure of distance, the photometry of stars can improve the distance inference.
The intrinsic temperatures and luminosities of stars tightly correlate leading to the observed colors and absolute magnitudes of stars also tightly correlating, usually visualized as the color--magnitude diagram (\cmd).
The tight correlation of the \cmd\ benefits our inference because the absolute magnitude of a star contains distance information.
We include this distance information in the inference of the true parallax by using the \cmd\ as our prior.
Specifically, we use the \cmd\ generated from the data themselves as our prior.
Instead of assuming a functional form for the spatial distribution of stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar models (\citealt{gaia16}), we build a noise-deconvolved estimate of the \cmd\ from the observed colors and absolute magnitudes of our full set of stars.
This \cmd\ model becomes, for each star---given its observed color and magnitude (and uncertainties on those)---a prior pdf for the distance
to that star.

Because our prior in the \cmd\ space, our inference will include not only the parallax from \gaia, but also the photometry of the stars. For inferring the true parallax, including generating our \cmd\ prior, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$. Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\mathbf{y}$ contains a transformation of the dust corrected absolute magnitude

@@

The transformation of the dust corrected absolute magnitude, Equation \ref{eq:transform}, is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain. Gaussian uncertainties are required to generate our prior using \xd\, but more about this and our dust corrections below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the Greene dust map

\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\mathbf{Y}_n = [10^{0.2\,M^{true}_{J,n} + 2}, (J-K)^{true}_n] \\
\mathrm{data \, vector} \quad &\mathbf{y}_n = [\varpi_n\,10^{0.2J^c_n}, (J- K)^c_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^c_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J - K)^c_n = J_n - K_n - Q_{JK}\,E(B-V)_n \\
\mathrm{assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}

@@

If life were simple, our inference would be a 1D problem using Bayes theorem looking something like:
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.

But life isn't simple, and our prior is a 2D function in color and transformed absolute magnitude space, so our full inference in done in 2D space as well.
We assume (as \gaia\ suggests) that the parallax likelihood is Gaussian, as well as the $(J-K)^c_n$ color likelihood. So our full likelihood is
\begin{equation}
p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) = \mathcal{N}(\mathbf{y}_n \given \mathbf{Y}_n, V_n)
\end{equation}

with \\
$V_n = $
\[
\begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & (\sigma_{\varpi}\,10^{0.2\,J_n^C})^2
\end{bmatrix}
\]

where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallas uncertainty dominates over the $J$ band uncertainty, so we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty. This is to keep our uncertainties Gaussian so we can use \xd\ generate our prior below.
This is the probability of the data vector given the true vector, but we
want to flip it around and infer the true vector and therefore the true parallax, or distance. We do this
analogous to Equation \ref{eq:bayes}.

\begin{equation}
p(\mathbf{Y}_n \given \mathbf{y}_n, I_n) = p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) \, p(\mathbf{Y}_n)
\end{equation}

where $p(\mathbf{Y}_n)$ is our prior.

To generate our prior, we fit the data in the slightly transformed \cmd\ space, Equation \ref{eq:transform}, using \xd\ which deconvolves the data with its uncertainties, which are heterogeneous measurement uncertainties. It generates the underlying distribution from which the uncertain data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the marginalized likelihood of the data, marginalized over the true data.

@@

\begin{eqnarray}
\alpha &=& \left\{\mathrm{amp}_k, \mu_k, V_k\right\}_{k=1}^K \; k \in \mathrm{Gaussians} \nonumber\\
       &=& \argmax_{\alpha} \, \Pi_n \, p(\mathbf{y}_n \given \alpha) \nonumber\\
       &=& \argmax_{\alpha} \, \Pi_n \, \int p(\mathbf{y}_n \given \mathbf{Y}_n) \, p(\mathbf{Y}_n \given \alpha)\mathrm{d\mathbf{Y}_n}
\label{eq:xdmml}
\end{eqnarray}

To infer $\varpi_{\true}$ we first do the inference in the 2D space of the \cmd

\begin{equation}
p(\mathbf{Y} \given \mathbf{y}, \alpha) = \frac{1}{Z} \, p(\mathbf{y} \given \mathbf{Y}) \, p(\mathbf{Y} \given \alpha)
\label{eq:posterior}
\end{equation}

We then project the 2D posterior into the 1D transformed absolute magnitude space, and do a little algebra to get

\begin{eqnarray}
p(\mathbf{Y}_0 \given \mathbf{y}, \alpha) &\rightarrow& p(\varpi_{\true} \given y, \alpha) \, f(\varpi_{\true}) \\
f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}
with $f(\varpi_{true})$ being a window function to insure the true parallax is positive and put it on a similar grid for all the posteriors.
And with a jacobian this becomes a posterior on distance.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{priorNgaussComparison.png}
\caption{Prior: the XD generated mixture of gaussians we use as our prior. To determine the number of gaussians we used cross validation, using 70\% of our sample as a training set and 30\% as the test set. Increasing the number of gaussians we used continued to increase the likelihood of the test set of data, showing that we are no over fitting. Using more gaussians as our prior did not substantially change our posterior results, but it did start to highlight the features of the CMD.}
\label{fig:priorCompare}
\end{figure}

@@

Some discussion of the inconsistency of how we treat dust within the probabilistic framework.
